A decentralized network support middleware for MMOGs

### 2009-09-05 textos/PlanoDoutorado/planodoc_lugano.tex

%
% Informações gerais
%
\title{Um modelo de suporte distribuído para jogos MMOG em cenários com recursos computacionais limitados%Uma proposta de arquitetura híbrida (par-a-par e cliente-servidor) para jogos online maciçamente distribuídos 
}

@@

This graduate school study plan proposes a support model for MMOGs (massively multiplayer online games) by using systems similar to those created for large-scale sharing of computing resources, such as peer-to-peer networks. Following this model, a system could be formed on a peer-to-peer network to act as a server for some MMOG. Then, a portion of the game's communication and processing load would then be delegated to each node of the server system -- more precisely, the game's virtual environment would be partitioned into regions, each of which managed by one or more server nodes. Such a system would be perceived by the players as a central to which they connect via a \emph{gateway}. Each player is then seamlessly redirected to the appropriate server node. However, for such server system to work as expected, some issues must be addressed, such as maintaining the game state consistend among the different game server nodes, so each player has the same vision; management and balancing of the load assigned to each server node; what to do when a node crashes or leaves the server system and, finally, how to broadcast and store the game state to the players in a distributed manner in a way such that its most recent state can be restored even in the presence of system failures.

@@

Massively multiplayer online games (MMOGs) are the core theme of this study plan. Usually a client-server infrastructure is used to support this kind of games. By following this plan, a distribution model will be proposed, reducing the maintenance cost of the game server. In the next sections, the context of multiplayer games in general and MMOGs will be presented.

@@

In the past few decades, since the creation of the personal computer, computer games have enjoyed an evergrowing popularity. After the acess to the Internet became widely available to the general public, with increasingly fast and cheap domestic connections, multiplayer online games has become an important market. To illustrate this, a study from 2006 estimated that the online games market would grow from US\$ 3.4 billion in 2005 to US\$ 13 billion in 2011 \cite{intelligence2006online}. The main characteristic of the MMOGs is the large number of concurrent players interacting with one another through the Internet. This kind of game may be divided into several categories, such as:

@@

\item RTS: \emph{real-time strategy}, where each player commands an army, similarly to a chess game, except that the game play is not divided in turns, i.e. it's a real-time game and the player who thinks faster than the other may have an advantage. Nevertheless, the action is usually slower than in a FPS game;

\item RPG: \emph{role-playing game}. In this kind of games, each player has a characters who evolves with time, acquiring power and accumulating wealth. These games are characterized by not having a predetermined end of the ``match'', creating a virtual world with persistent state. The player may, then, connect to this world, play for some time with his character, disconnect and, later, reconnect and continue from the point he was when he disconnected for the last time, for his state is stored in the game server.

@@

After single-player computer games evolved to multiplayer online games with the advent of the Internet, MMOGs emerged as a new trend because of the decrease of the domestic Internet connection cost and the increase of its average bandwidth. These kind of games have become very popular in the latest years because they can not only be played online, as they also allow the interaction of a massive number of players at the same time. In the most succesful cases, such as World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and EVE Online \cite{eveonline}, for example, tens of thousands of players are supported simultaneously \cite{chen2006pch}. These players can, then, interact with one another and with the virtual environment of the game.

One of the main features of these games, present in almost all commercial MMOGs, is that the matches are very long. In some MMORTS (real-time strategy MMOGs), such as Travian \cite{travian}, the games come to last from six months to one year. MMORPGs (MMOGs in RPG style), as is the case of World of Warcraft, the games has no given time to end. The worlds of these games are so vast and populous -- as regards the number of players -- that they may evolve and modify their state independently of some players being playing it or not, because there is always someone connected so the games never stops. For this reason, it is expected that each player can disconnect and continue playing later, without having to start from scratch. Moreover, it is expected that changes made on the game world persist, no matter how many times each player enters and leaves.

Generally, each player controls an entity called \emph{avatar}, which is nothing more than his representation in the virtual environment -- a character who executes the orders given by the player, therefore interfering in the history and in the outcome of the game. For the game to remain consistent for the different participants, each action performed by each avatar must be processed by the server, which calculates the game state resulting from such action. This new state is then broadcast to other players. Other changes in the game world, such as destruction or creation of objects in the environment, and events such as actions of characters controlled by the server must also be sent to the players involved. It is then through these messages sent by the player to the server and in the opposite direction that that players can interact with each other and with the simulated world of the game.

However, it is perceived that such a mechanism can easily saturate the bandwidth and overload the cpu(s) of the server. What is usually done is build an infrastructure with a large central server, usually a cluster with its nodes connected through local high-speed and low latency network, with an Internet connection of some hundreds or thousands of MBps. The greatest cost comes from maintaining this connection, whose main purpose is to maintain every player as up-to-date as possible about the game state -- naturally, tolerating some delay, which should not be over a given threshold.

@@

As mentioned in the previous section, MMOGs are characterized by a large number of players. To support this great number of participants, a client-server infrastructure is generally used, and all interaction is done through the server machine. In this paradigm, the server computer has the official copy of the game state and it is responsible for computing its change over time, while the game client is responsible for presenting to the user the game state sent by the server. The server is also responsible for updating the clients, in real time, on the changes that occur in the virtual environment of the game. Because of the high number of participants, the centralization on the game server makes the maintanance cost very high. Generally, it is required a connection to the Internet with several GBps of bandwidth and enough processing power to run the simulation of the virtual world \cite{feng2007wnn}. It becomes impossible, for example, the maintenance of massively multiplayer online games by small or independent groups on a limited budget.

Because of this high cost, alternative solutions have been sought in order to distribute somehow the communication and processing loads originated from the game. Generally, it is proposed the use of peer-to-peer networks -- or even computer GRIDS \cite{butterfly, wang2004gem}. However, an MMOG is a simulation of a dynamic virtual environment, with a maximum tolerable delay between sending the player's actions and receiving the resulting state. MMOGs have also other requirements that must be observed carefully so that the players may have a good game experience. To distribute the network support structure among different machines connected via the Internet -- which is the case of the peer-to-peer networks -- while meeting the requirements of an MMOG becomes a challenging task that has motivated many researchers in recent years.

@@

Based on the current state of the art and in previous work -- such as in \cite{bezerra2009lbs, bezerra2008a3, vilanova2008psm, cecin2006pps, cecin2004fsa} --, we can summarize a few requirements that we consider critical for the game network support infrastructure to be considered adequate, and which will be dealt with in the doctoral course:

@@

	\item \textbf{Availability} - In general, MMOG games are not free and, because of that, their users tend to be demanding when it comes to availability. Thus, it is desired the system is always available;

	\item \textbf{Consistency} - Multiplayer Games consist of a virtual environment shared among the players participating in the same game. It is therefore necessary that the environment is represented in the same way on the machines of the various players so that they can interact with each other and with the world in an appropriate manner;

@@

	\item \textbf{Scalability} - Taking into account the number of participants in an MMOG game, it is necessary that the system that supports it is able to maintain the quality of service despite the increasing burden of processing and communication required for this.

@@

Another approach is the \textbf{peer-to-peer architecture}, where the simulation is divided among the computers involved. There can be a system without any server, where the peers, which are the machines of the players, come into some kind of agreement for the various steps of the simulation. Regarding scalability, this approach is not optimal, since ensuring such "agreement"{} is costly in terms of message exchange \cite{lamport1982bgp}. Even if one of the peers is elected to decide the course of the simulation, still persists the problem of all peers needing to exchange messages with one another. Having $n$ peers, there is a complexity of O({$n^2$}) message exchanges for each step of the simulation. Clearly, this approach is not as scalable as should be a system to support a massively multiplayer online game. Moreover, it would be necessary to distribute the storage and retrieval of the states of the game.

@@

It is intended, during graduate school, research and develop new solutions to these problems, seeking those that are most appropriate in a context of massively multiplayer online games. With these issues resolved, it may be provided to MMOGs availability and fault tolerance -- due to the replication of server nodes in the same region -- and scalability, while maintaining the communication delay within an acceptable range for the game and the a lower financial cost, compared to traditional MMOG servers.

@@

The main objective of this doctoral plan is to create a decentralized architectural model to support massively multiplayer online games. Following this model, it should be possible to form a system of geographically distributed server nodes, using the Internet as a means of connection between them. This distribution should be transparent to the players, to whom the system will be seen as a single MMOG server, which maintains the game world consistent, available and with a good quality of service.

@@

\item Design techniques and algorithms that aim to reduce intra-server and server-client communication, in order to save resources of the server nodes, reduce delays in the communication between them and expand the audience of the game due to the reduction in the bandwidth required to play it and, besides, reducing the requirements for a computer to be part of the server system, without undermining the game experience of the players.

@@

>>>>

%
% Informações gerais
%
\title{A decentralized network support middleware for MMOGs% with limited computer resourcesem cenários com recursos computacionais limitados%Uma proposta de arquitetura híbrida (par-a-par e cliente-servidor) para jogos online maciçamente distribuídos 
}

@@

This doctorate study plan proposes the research and creation of a novel support model for MMOGs (massively multiplayer online games), by using large-scale systems similar to those created for sharing of computing resources, such as file sharing peer-to-peer networks. Following this model, a distributed system could be formed to act as a server for an MMOG. Then, a portion of the game's communication and processing load would be delegated to each node of the server system -- more precisely, the game's virtual environment would be partitioned into regions, each of which managed by one or more server nodes. Such a system would be perceived by the players as an unique central servel to which they connect via a gateway. Upon connected, each player is seamlessly redirected to the appropriate server node. However, for such server system to work as expected, some issues must be addressed, such as maintaining the game state consistent among the different game server nodes, so each player has the same vision; management and balancing of the load assigned to the different server nodes; what to do when a node crashes or leaves the server system and, finally, how to broadcast and store the game state to the players in a distributed manner so that its most recent state can be restored even in the presence of system failures.

@@

Massively multiplayer online games (MMOGs) are the theme of this study plan. Usually a client-server infrastructure is used to support this kind of games. This plan proposes the investigation and definition of a novel distribution model to be used for MMOGs, reducing the maintenance cost of the game server. In the next sections, the state of the art regarding this area will be presented.

@@

In the past few decades, since the creation of the personal computer, computer games have enjoyed an evergrowing popularity. After the acess to the Internet became widely available to the general public, with increasingly fast and cheap domestic connections, multiplayer online games have become an important market. To illustrate this, a study from 2006 estimated that the online games market would grow from US\$ 3.4 billion in 2005 to US\$ 13 billion in 2011 \cite{intelligence2006online}. The main characteristic of the MMOGs is the large number of concurrent players interacting with one another through the Internet. This kind of game may be divided into several categories, such as:

@@

\item RTS: \emph{real-time strategy}, where each player commands an army, similarly to a chess game, except that the game play is not divided in turns, i.e. it's a real-time game and the player who thinks faster than the others may have an advantage. Nevertheless, the action is usually slower than in an FPS game;

\item RPG: \emph{role-playing game}. In this kind of game, each player has a characters who evolves with time, acquiring power and accumulating wealth. These games are characterized by not having a predetermined end of the ``match'', creating a virtual world with persistent state. The player may, then, connect to this world, play for some time with his character, disconnect and, later, reconnect and continue from the point he was when he disconnected for the last time, for his state is stored in the game server.

@@

After single-player computer games evolved to multiplayer online games with the advent of the Internet, MMOGs emerged as a new trend because of the decrease of the domestic Internet connection cost and the increase of its average bandwidth. This kind of games have become very popular in the latest years because they allow the online interaction of a massive number of players at the same time. In the most succesful cases, such as World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and EVE Online \cite{eveonline}, for example, tens of thousands of players are supported simultaneously \cite{chen2006pch}. These players can, then, interact with one another and with the virtual environment of the game.

One of the main features of these games, present in almost all commercial MMOGs, is that the matches are very long. In some MMORTS (real-time strategy MMOGs), such as Travian \cite{travian}, the games come to last from six months to one year. MMORPGs (MMOGs in RPG style), as is the case of World of Warcraft, the games have no given time to end. The worlds of these games are so vast and populous -- as regards the number of players -- that they may evolve and modify their state independently of some players being playing it or not, because there is always someone connected so the game never stops. For this reason, it is expected that each player can disconnect and continue playing later, without having to start from scratch. Moreover, it is expected that changes made on the game world persist, no matter how many times each player enters and leaves.

Generally, each player controls an entity called \emph{avatar}, which is nothing more than his representation in the virtual environment -- a character who executes the orders given by the player, therefore interfering in the history and in the outcome of the game. For the game to remain consistent for the different participants, each action performed by each avatar must be processed by the server, which calculates the game state resulting from such action. This new state is then broadcast to other players. Other changes in the game world, such as destruction or creation of objects in the environment, and events such as actions of characters controlled by the server must also be sent to the players involved. It is then through these messages sent by the player to the server and in the opposite direction that players can interact with each other and with the simulated world of the game.

However, it is perceived that such a mechanism can easily saturate the bandwidth and overload the cpu(s) of the server. What is usually done is build an infrastructure with a large central server, usually a cluster with its nodes connected through a local high-speed and low latency network, with an Internet connection of some hundreds or thousands of MBps. The greatest cost comes from maintaining this connection, whose main purpose is to maintain every player as up-to-date as possible about the game state -- naturally, tolerating some delay, which should not be over a given threshold.

@@

As already mentioned, MMOGs are characterized by a large number of players. To support this great number of participants, a client-server infrastructure is generally used, and all interaction is done through the server machine. In this paradigm, the server computer has the official copy of the game state and it is responsible for computing its change over time, while the game client is responsible for presenting to the user the game state sent by the server. The server is also responsible for updating the clients, in real time, on the changes that occur in the virtual environment of the game. Because of the high number of participants, the centralization on the game server makes the maintanance cost very high. Generally, it is required that the MMOG server has a connection to the Internet with several GBps of bandwidth and enough processing power to run the simulation of the virtual world \cite{feng2007wnn}. It becomes impossible, for example, the maintenance of massively multiplayer online games by small or independent groups on a limited budget.

Because of this high cost, alternative solutions have been sought in order to distribute somehow the communication and processing loads originated from the game. Generally, it is proposed the use of peer-to-peer networks -- or even computer GRIDS \cite{butterfly, wang2004gem}. However, an MMOG is a simulation of a dynamic virtual environment, with a maximum tolerable delay between sending the player's actions and receiving the resulting state. MMOGs have also other requirements that must be observed carefully so that the players may have a good game experience. For example, as in a fully decentralized peer-to-peer MMOG the peers may have to calculate themselves the outcome of the players' actions, a malicious player could try to hack the software running in his computer in order to forge game states that provide him unfair advantages over the other players. To distribute the network support structure among different machines connected via the Internet -- which is the case of the peer-to-peer networks -- while meeting the requirements of an MMOG becomes a challenging task that has motivated many researchers in recent years.

@@

Based on the current state of the art and in previous work -- such as in \cite{bezerra2009lbs, bezerra2008a3, vilanova2008psm, cecin2006pps, cecin2004fsa} --, we can summarize a few requirements that we find critical for the game network support infrastructure to be considered adequate, and which will be dealt with in the doctoral course:

@@

	\item \textbf{Availability} - In general, MMOGs require that the players make some kind of payment for account maintenance and, because of that, they tend to be demanding when it comes to availability. Thus, it is desired the system is always available;

	\item \textbf{Consistency} - Multiplayer games consist of a virtual environment shared among the players participating in the same match. It is therefore necessary that the environment is represented in the same way on the machines of the various players so that they can interact with each other and with the world in an appropriate manner;

@@

	\item \textbf{Scalability} - Taking into account the number of participants in an MMOG, it is necessary that the system that supports it is able to maintain the quality of service despite the increasing processing and communication load required for this.

@@

Another approach is the \textbf{peer-to-peer architecture}, where the simulation is divided among the computers involved. There can be a system without any server, where the peers, which are the machines of the players, come into some kind of agreement for the various steps of the simulation, calculating the game state resulting of each action of the players. Regarding scalability, this approach is not optimal, since ensuring such "agreement"{} is costly in terms of message exchange \cite{lamport1982bgp}. Even if one of the peers is elected to decide the course of the simulation, still persists the problem of all peers needing to exchange messages with one another. Having $n$ peers, there is a complexity of O({$n^2$}) message exchanges for each step of the simulation. Clearly, this approach is not as scalable as should be a system to support a massively multiplayer online game. Moreover, it would be necessary to distribute the storage and retrieval of the states of the game.

@@

It is intended, during the doctoral course, to research and develop new solutions to these problems, seeking those that are most appropriate in a context of massively multiplayer online games. With these issues resolved, it may be provided to MMOGs availability and fault tolerance -- due to the replication of server nodes in the same region -- and scalability, while maintaining the communication delay within an acceptable range for the game and the a lower financial cost, compared to traditional MMOG servers.

These solutions will be wrapped together in a middleware, which will provide them as services. These services will be accessible via an API (application programming interface) to allow an easier development of a massively multiplayer game. The game programmer will be able to use these services without needing to have a deep understanding of the distribution model used by the middleware. To the application developer, a high-level interface will be provided, while the middleware will take care of the tasks performed in lower levels. The communication abstraction provided to the game developer may be: publish-subscribe, as in PADRES \cite{fidler2005padres}, distributed shared memory, as in Sinfonia \cite{aguilera2007sinfonia}, message passing or any other that best suits the development of an MMOG.

@@

The main objective of this doctoral plan is to create a decentralized architectural model to support massively multiplayer online games. Following this model, it should be possible to form a system of geographically distributed server nodes, using the Internet as a means of connection between them. This distribution should be transparent to the players, to whom the system will be seen as a single MMOG server, which maintains the game world consistent, available and with a good quality of service. As a final product, a middleware for MMOG development with a high-level interface will be created.

@@

\item Design techniques and algorithms that aim to reduce intra-server and server-client communication, in order to save resources of the server nodes, reduce delays in the communication between them and expand the audience of the game due to the reduction in the bandwidth required to play it and, besides, reducing the requirements for a computer to be part of the server system, without undermining the game experience of the players;

\item Create a middleware containing the desired services, along with an high-level API to make them accessible to the game developer.

@@

\section{High-level application programming interface}

In order to offer to application (game) developers an easy-to-use set of services, a high-level application programming interface must be created for Cosmmus. This set of services will compose the Cosmmus middleware, which will encapsulate all the low level operations regarding the distribution of the game. There are some options to define this distribution model -- and, also, the abstraction provided to the game developer --, like distributed shared memory, publish/subscribe, message passing and others. Each one of these presents both advantages and disadvantages, but the publish/subscribe seems to be the most suited for the development of MMOGs, since the entities (avatars and objects) present in the game's virtual world imply state update messages, in which every player may be interested or not.
